---
title: "Problem 2"
author: "Sharon Njeri"
date: "2024-03-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(forecast)
library(readxl)
```
# Load the data
```{r}
# Load the data
data <- read_excel("C:/Users/User/Downloads/Makert Prices 2022.xlsx")
library(readxl)
```

```{r}
# merging entries with different case
data <- data %>%
  mutate(Seasons = case_when(
    Seasons == "dry" ~ "Dry",
    Seasons == "wet" ~ "Wet",
    TRUE ~ Seasons
  ))
```

```{r}
# Convert the Dates column to Date type
data$Dates <- as.Date(data$Dates, format="%m/%d/%Y")
```

```{r}
# Perform exploratory analysis
summary(data)
```

```{r}
# Create dummy variables for seasons
data$Drought <- ifelse(data$Seasons == "Drought", 1, 0)
data$Wet <- ifelse(data$Seasons == "Wet", 1, 0)
data$Dry <- ifelse(data$Seasons == "Dry", 1, 0)
```

# Modelling
```{r}
# creating a matrix of external regressors for the arima model
exreg= as.matrix(cbind(data$Drought,data$Wet,data$Dry))
colnames(exreg) <- c("Drought", "Wet", "Dry")
```

```{r}
# Fit ARIMA model with  season as external regressor.
Bull_model<- auto.arima(data$Bull, xreg = exreg )
Cow_model<- auto.arima(data$Cow, xreg = exreg )
Heifer_model<- auto.arima(data$Heifer, xreg = exreg )
Steer_model<- auto.arima(data$Steer, xreg = exreg )
```

```{r}
# Get the summary of each model
summary(Bull_model)
```
The model suggests that selling during a drought could lead to a decrease in price (coefficient: -1441.386), while selling during wet (coefficient: 1300.870) and dry (coefficient: 997.2451) seasons could lead to an increase in price. However, the standard errors are quite large, indicating a high degree of uncertainty in these estimates.
```{r}
summary(Cow_model)
```
The model suggests that selling during a drought could lead to a slight decrease in price (coefficient: -34.9539), while selling during wet (coefficient: 1309.246) and dry (coefficient: 94.7516) seasons could lead to an increase in price. The standard errors are also quite large for this model.
```{r}
summary(Heifer_model)
```
Cow: The model suggests that selling during a drought could lead to a slight decrease in price (coefficient: -34.9539), while selling during wet (coefficient: 1309.246) and dry (coefficient: 94.7516) seasons could lead to an increase in price. The standard errors are also quite large for this model.
```{r}
summary(Steer_model)
```

```{r}

# Get the seasons for the last 12 months
last_12_seasons <- tail(exreg, 12)

# Create a future exreg matrix based on these seasons
future_exreg <- model.matrix(~last_12_seasons-1)
colnames(future_exreg) <- c("Drought", "Wet", "Dry")

# Forecast the next 12 months
Bull_forecast <- forecast(Bull_model, h=12, xreg=future_exreg)
Cow_forecast <- forecast(Cow_model, h=12, xreg=future_exreg)
Heifer_forecast <- forecast(Heifer_model, h=12, xreg=future_exreg)
Steer_forecast <- forecast(Steer_model, h=12, xreg=future_exreg)

```

# Plotting the forecasts.

1.  Bull

```{r,fig.dpi=900}
# Create a data frame for the historical data
historical_data <- data.frame(
  Dates = data$Dates,
  Price = data$Bull
)

# Create a data frame for the forecasted data
forecast_data <- data.frame(
  Dates = seq(max(data$Dates), by = "month", length.out = 12),
  Price = as.numeric(Bull_forecast$mean)
)

# Combine the historical and forecasted data
combined_data <- rbind(historical_data, forecast_data)

# Convert 'Dates' to Date class
combined_data$Dates <- as.Date(combined_data$Dates)

ggplot() +
  geom_line(data = historical_data, aes(x=Dates, y=Price), color = "blue") +
  geom_line(data = forecast_data, aes(x=Dates, y=Price), color = "red") +
  labs(x = "Dates", y = "Price", title = "Market Price Over Time: Bull")
```



2.  Cow

```{r,fig.dpi=900}

# Create a data frame for the historical data
historical_data_cow <- data.frame(
  Dates = data$Dates,
  Price = data$Cow
)

# Create a data frame for the forecasted data
forecast_data_cow <- data.frame(
  Dates = seq(max(data$Dates), by = "month", length.out = 12),
  Price = as.numeric(Cow_forecast$mean)
)

# Combine the historical and forecasted data
combined_data_cow <- rbind(historical_data_cow, forecast_data_cow)

# Convert 'Dates' to Date class
combined_data_cow$Dates <- as.Date(combined_data_cow$Dates)

ggplot() +
  geom_line(data = historical_data_cow, aes(x=Dates, y=Price), color = "blue") +
  geom_line(data = forecast_data_cow, aes(x=Dates, y=Price), color = "red") +
  labs(x = "Dates", y = "Price", title = "Market Price Over Time: Cow")

```

3.  Heifer
```{r, fig.dpi=900}
# Create a data frame for the historical data
historical_data_heifer <- data.frame(
  Dates = data$Dates,
  Price = data$Heifer
)

# Create a data frame for the forecasted data
forecast_data_heifer <- data.frame(
  Dates = seq(max(data$Dates), by = "month", length.out = 12),
  Price = as.numeric(Heifer_forecast$mean)
)

# Combine the historical and forecasted data
combined_data_heifer <- rbind(historical_data_heifer, forecast_data_heifer)

# Convert 'Dates' to Date class
combined_data_heifer$Dates <- as.Date(combined_data_heifer$Dates)

ggplot() +
  geom_line(data = historical_data_heifer, aes(x=Dates, y=Price), color = "blue") +
  geom_line(data = forecast_data_heifer, aes(x=Dates, y=Price), color = "red") +
  labs(x = "Dates", y = "Price", title = "Market Price Over Time: Heifer")
```
4.  Steer
```{r,fig.dpi=900}
# Steer
# Create a data frame for the historical data
historical_data_steer <- data.frame(
  Dates = data$Dates,
  Price = data$Steer
)

# Create a data frame for the forecasted data
forecast_data_steer <- data.frame(
  Dates = seq(max(data$Dates), by = "month", length.out = 12),
  Price = as.numeric(Steer_forecast$mean)
)

# Combine the historical and forecasted data
combined_data_steer <- rbind(historical_data_steer, forecast_data_steer)

# Convert 'Dates' to Date class
combined_data_steer$Dates <- as.Date(combined_data_steer$Dates)

ggplot() +
  geom_line(data = historical_data_steer, aes(x=Dates, y=Price), color = "blue") +
  geom_line(data = forecast_data_steer, aes(x=Dates, y=Price), color = "red") +
  labs(x = "Dates", y = "Price", title = "Market Price Over Time: Steer")

```
